---
title: "The Illinois People Over Parking Act"
author: "Michael McLean - Member, Abundant Housing Illinois"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output_file: "docs/index.html"
output:
  html_document:
    theme: flatly
    toc: false
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

Note: The People Over Parking Act was made possible by the volunteers of [Abundant Housing Illinois](https://abundanthousingillinois.org/). Please consider joining as a member if you are interested in organizing for more land use reform in Illinois.

### What Are Parking Mandates?

Parking mandates, also known as parking minimums, are zoning regulations that require developers to build a minimum number of parking spaces with new housing and commercial developments. These mandates typically specify a certain number of parking spaces per unit of housing, square footage of retail space, or other metrics depending on the development type.

#### Why Parking Mandates Are Costly:

-   **Increased Housing Costs**: Each parking space can cost \$30,000-\$75,000 to build, significantly raising the cost of housing development and ultimately housing prices.
-   **Reduced Housing Supply**: Land used for parking cannot be used for additional housing units, limiting the overall housing supply.
-   **Environmental Impact**: Excessive parking encourages car dependency, increasing traffic congestion and carbon emissions.
-   **Inefficient Land Use**: Parking lots create "dead spaces" in urban areas that could otherwise be used for housing, businesses, or public spaces.
-   **Economic Burden**: Many parking spaces sit empty much of the time, representing wasted resources and opportunity costs.

#### Why Parking Mandates Should Be Repealed:

-   **Market-Based Solutions**: Developers can better determine the appropriate amount of parking based on actual demand rather than arbitrary requirements.
-   **Transit-Oriented Development**: Eliminating parking mandates near transit encourages development that leverages existing public transportation infrastructure.
-   **Affordability**: Reducing or eliminating parking requirements can make housing more affordable and accessible.
-   **Sustainability**: Less parking promotes walking, cycling, and public transit use, reducing carbon emissions.
-   **Vibrant Communities**: Space previously dedicated to parking can be repurposed for housing, businesses, and community amenities.

### Enter: The People Over Parking Act

```{r packages, results='hide'}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

# Install packages if not already installed
required_packages <- c("tidyverse", "sf", "lwgeom", "leaflet", "leaflet.extras",
                       "data.table", "zip", "httr", "lubridate", "mapview", "tigris", "kableExtra", "tidytransit")

new_packages <- required_packages[!required_packages %in% installed.packages()[,"Package"]]
if(length(new_packages)) install.packages(new_packages)

# Load required packages
library(tidyverse)
library(sf)
library(leaflet)
library(lwgeom)
library(leaflet.extras)
library(data.table)
library(zip)
library(httr)
library(lubridate)
library(mapview)
library(tigris) # For Illinois state boundary
library(kableExtra) # Added for enhanced table formatting
library(tidytransit) # For GTFS data processing

# Disable s2 processing to avoid geometry validation issues
sf_use_s2(FALSE)

# Load Illinois state boundary for clipping transit areas
illinois_boundary <- states(cb = TRUE, year = 2023) %>%
  filter(STUSPS == "IL") %>%
  st_transform(4326)  # Transform to WGS84
```

```{r load_existing_data, results='hide'}
## Load Existing Hub Data and Corridor Data

# Load the existing hub processing results (from the original analysis)
# We'll reuse the hub identification logic from the original script

# Load GTFS utility functions from extracted modules
source("R/agency_metadata.R")  # Centralized agency configuration
source("R/tidytransit_integration.R")  # tidytransit wrapper functions
source("R/gtfs_download.R")
source("R/gtfs_normalize.R")
source("R/gtfs_validate.R")
source("R/gtfs_processing.R")
source("R/spatial_validate.R")
source("R/spatial_clustering.R")
source("R/frequency_calc.R")
source("R/hub_identification.R")
source("R/hub_processing.R")
source("R/corridor_processing.R")
source("R/buffer_processing.R")
source("R/map_creation.R")
source("R/summary_stats.R")
```

```{r download_gtfs, results='hide'}
## Download and Process GTFS Data

# Get agency configurations from centralized metadata
# Now includes all 14 Illinois transit agencies (6 existing + 8 new)
agency_configs <- get_agency_configs_for_download()

# Process all GTFS data using tidytransit: download, validate, add unique IDs
agencies_data <- process_agencies_with_tidytransit(agency_configs, validate = TRUE)

# Combine data from all agencies into unified tables
combined_tables <- combine_agency_data(agencies_data)

# Extract combined tables for easy access
all_stops <- combined_tables$stops
all_routes <- combined_tables$routes
all_trips <- combined_tables$trips
all_stop_times <- combined_tables$stop_times
all_calendar <- combined_tables$calendar
all_calendar_dates <- combined_tables$calendar_dates
all_shapes <- combined_tables$shapes

# Enrich stop_times with route information (adds unique_route_id and route_type)
cat("Enriching stop_times with route information...\n")
all_stop_times <- enrich_stop_times(all_stop_times, all_trips, all_routes)

# Individual agency data for backward compatibility
# Access via agencies_data[[agency_id]]$data
cta_data <- agencies_data$cta$data
pace_data <- agencies_data$pace$data
metra_data <- agencies_data$metra$data
metro_stl_data <- agencies_data$metro_stl$data
cumtd_data <- agencies_data$cumtd$data
rmtd_data <- agencies_data$rmtd$data
# New agencies (8):
metrolink_quad_cities_data <- agencies_data$metrolink_quad_cities$data
citylink_data <- agencies_data$citylink$data
smtd_data <- agencies_data$smtd$data
dekalb_data <- agencies_data$dekalb$data
connect_transit_data <- agencies_data$connect_transit$data
dpts_data <- agencies_data$dpts$data
galesburg_data <- agencies_data$galesburg$data
gowest_data <- agencies_data$gowest$data

# Store validation results for reference
validation_results <- lapply(agencies_data, function(a) a$validation)
```

```{r explore_direction_data, results='hide'}
## Explore direction_id availability in GTFS data

# Check if direction_id exists in trips data
cat("=== Checking direction_id availability ===\n\n")

# Iterate over all agencies dynamically
all_agency_ids <- get_all_agency_ids()
for (agency_id in all_agency_ids) {
  trips_data <- agencies_data[[agency_id]]$data$trips
  agency_name <- get_agency_display_name(agency_id)

  if ("direction_id" %in% names(trips_data)) {
    cat(sprintf("%s: Has direction_id field\n", agency_name))
    cat(sprintf("  - Total trips: %d\n", nrow(trips_data)))
    cat(sprintf("  - Trips with direction_id: %d\n", sum(!is.na(trips_data$direction_id))))
    cat(sprintf("  - Unique direction values: %s\n",
                paste(sort(unique(trips_data$direction_id)), collapse = ", ")))
  } else {
    cat(sprintf("%s: NO direction_id field\n", agency_name))
  }
  cat("\n")
}

# If direction_id exists in any agency, ensure it exists in all and rebuild all_trips
has_direction_id <- any(sapply(all_agency_ids, function(agency_id) {
  "direction_id" %in% names(agencies_data[[agency_id]]$data$trips)
}))

if (has_direction_id) {
  # Ensure direction_id exists in all agency datasets (add NA if missing)
  for (agency_id in all_agency_ids) {
    if (!"direction_id" %in% names(agencies_data[[agency_id]]$data$trips)) {
      # Add direction_id column using regular assignment (works with data.table)
      agencies_data[[agency_id]]$data$trips$direction_id <- NA_integer_
    }
  }

  # Rebuild all_trips with direction_id from all agencies dynamically
  all_trips_list <- lapply(all_agency_ids, function(agency_id) {
    agencies_data[[agency_id]]$data$trips
  })
  all_trips <- rbindlist(all_trips_list, fill = TRUE)

  cat("direction_id added to all_trips\n")
  cat(sprintf("Trips with direction_id: %d / %d (%.1f%%)\n",
              sum(!is.na(all_trips$direction_id)),
              nrow(all_trips),
              100 * sum(!is.na(all_trips$direction_id)) / nrow(all_trips)))
}
```

```{r process_hubs_UPDATED, results='hide'}
## Identify Public Transportation Hubs

# Identify all transit hubs (rail stations + qualifying bus hubs)
# Now uses representative weekday service selection per agency
hub_results <- identify_all_hubs(
  all_stops, all_routes, all_trips,
  all_stop_times, all_calendar, all_calendar_dates
)

# Extract results
all_hubs_sf <- hub_results$all_hubs_sf
am_peak_bus_stops <- hub_results$am_peak_bus_stops
pm_peak_bus_stops <- hub_results$pm_peak_bus_stops

# Filter hubs to only those within Illinois boundary
# This excludes St. Louis metro stations in Missouri
all_hubs_sf <- st_intersection(all_hubs_sf, illinois_boundary)
cat(sprintf("Hubs after filtering to Illinois boundary: %d\n", nrow(all_hubs_sf)))
```

```{r process_corridors_and_buffers, results='hide'}
## Identify Corridors and Create Buffers

# Identify qualifying transit corridors using edge-based method
# This method automatically handles:
# 1. Single routes with ≤15 min frequency (their edges qualify)
# 2. Multiple routes combining to ≤15 min frequency (shared edge qualifies)
# by counting all trips on each corridor edge regardless of route
corridor_results <- identify_qualifying_corridors(
  all_stops,
  hub_results$am_peak_bus_stops,
  hub_results$pm_peak_bus_stops,
  all_trips,
  all_shapes
)

qualifying_corridor_segments_sf <- corridor_results$qualifying_corridor_segments
corridor_qualification_summary <- corridor_results$qualification_summary

# Create corridor buffers (1/8 mile along actual route geometry from GTFS shapes)
all_corridors_union_wgs84 <- create_corridor_buffers(
  qualifying_corridor_segments_sf,
  illinois_boundary
)

# Create hub buffers (1/2 mile around hubs)
hub_buffers <- create_hub_buffers(all_hubs_sf, illinois_boundary)

# Extract hub buffer components for area calculations
all_hub_areas_wgs84 <- hub_buffers$all_hub_areas
half_mile_buffers_wgs84 <- hub_buffers$half_mile_buffers

# Legacy hub unions (backward compatibility)
cta_hubs_union <- hub_buffers$cta_hubs_union
pace_hubs_union <- hub_buffers$pace_hubs_union
metra_hubs_union <- hub_buffers$metra_hubs_union
metro_stl_hubs_union <- hub_buffers$metro_stl_hubs_union
cumtd_hubs_union <- hub_buffers$cumtd_hubs_union
rmtd_hubs_union <- hub_buffers$rmtd_hubs_union

# New agencies available via hub_buffers$per_agency_union[[agency_id]]

# Combine all affected areas (hubs + corridors)
all_affected_areas_combined <- create_combined_buffers(
  all_hub_areas_wgs84, all_corridors_union_wgs84, illinois_boundary
)
```

```{r calculate_areas, results='hide'}
## Calculate Areas

# Calculate areas in square miles
hub_area_sqft <- st_area(all_hub_areas_wgs84)
hub_area_sqmi <- units::set_units(hub_area_sqft, "mi^2")

corridor_area_sqft <- st_area(all_corridors_union_wgs84)
corridor_area_sqmi <- units::set_units(corridor_area_sqft, "mi^2")

combined_area_sqft <- st_area(all_affected_areas_combined)
combined_area_sqmi <- units::set_units(combined_area_sqft, "mi^2")

# Total area covered by analysis
# Chicago MSA (6 IL counties): Cook: 953.6, DuPage: 336.5, Kane: 524.2, Lake: 470, McHenry: 611, Will: 849.2
# St. Louis MSA (3 IL counties): St. Clair: 674.6, Madison: 741.5, Monroe: 398.2
# Champaign-Urbana: Champaign County: 1,008
# Rockford: Winnebago County: 519
chicago_il_msa_area_sqmi <- 953.6 + 336.5 + 524.2 + 470 + 611 + 849.2
stlouis_il_msa_area_sqmi <- 674.6 + 741.5 + 398.2

# Calculate summary statistics
total_rail_hubs <- nrow(all_hubs_sf[all_hubs_sf$type == "rail", ])
total_bus_hubs <- nrow(all_hubs_sf[all_hubs_sf$type == "bus_hub", ])
total_qualifying_segments <- nrow(qualifying_corridor_segments_sf)
```

```{r summary_table}
# Create summary table for display
summary_data <- data.frame(
  Category = c("Rail Stations", "Qualifying Bus Hubs", "Qualifying Corridor Segments",
               "Hub Buffer Area", "Corridor Buffer Area", "Total Affected Area"),
  Count = c(total_rail_hubs, total_bus_hubs, total_qualifying_segments,
            round(as.numeric(hub_area_sqmi), 1),
            round(as.numeric(corridor_area_sqmi), 1),
            round(as.numeric(combined_area_sqmi), 1)),
  Units = c("stations", "hub clusters", "edge segments",
            "sq mi", "sq mi", "sq mi")
)

kable(summary_data, caption = "Transit Hub and Corridor Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r create_map}
## Create Interactive Map

# Convert shapes to linestrings for map display
cat("Converting GTFS shapes to linestrings for visualization...\n")
all_shapes_sf <- convert_shapes_to_linestrings(all_shapes)

# Create the interactive map with all layers
# Now includes corridor segments, all routes, and bus stops for detailed visualization
final_map <- create_interactive_map(
  all_hubs_sf = all_hubs_sf,
  all_affected_areas = all_affected_areas_combined,
  hub_buffers = hub_buffers,
  all_corridors_union = all_corridors_union_wgs84,
  qualifying_corridor_segments_sf = qualifying_corridor_segments_sf,
  all_shapes_sf = all_shapes_sf,
  all_stops = all_stops,
  illinois_boundary = illinois_boundary
)

final_map
```

### Methodology

This analysis implements the following methodology to identify qualifying transit hubs and corridors:

**Representative Service Selection:**

-   For each transit agency, we select a **single representative weekday service_id** rather than combining all weekday services
-   This prevents inflating trip counts when agencies define the same schedule multiple times for different calendar dates
-   Selection strategy: Find services running on the next upcoming Tuesday (a typical weekday), or the most recent Tuesday if none upcoming
-   If multiple services run on the same day, the service with the most trips is selected
-   This ensures frequency calculations reflect actual service levels, not artifacts of how calendar data is structured

**Peak Period Definition:**

-   Morning peak: 7:00 AM - 9:00 AM (120 minutes)
-   Afternoon peak: 4:00 PM - 6:00 PM (120 minutes)
-   Service is analyzed separately for each period; locations qualify if they meet frequency thresholds in **either** AM **or** PM

**Stop Clustering (Bus Hubs Only):**

-   Bus stops within **150 feet** of each other are spatially clustered to represent physical intersections
-   This prevents overcounting stops at the same street corner and aligns with the statutory concept of "an intersection of 2 or more bus routes"
-   Clustering uses Illinois State Plane East projection (EPSG:3435) for accurate distance measurement

**Route Overlap Verification (Bus Hubs Only):**

-   After clustering, we verify that routes actually overlap at the same physical intersection or street
-   Stop names (e.g., "Jackson & Lotus", "Oak Ave & Davis St") are parsed to extract street names
-   A cluster qualifies only if **2 or more routes share at least one common street name**
-   This helps ensure routes truly intersect at the location, not just have nearby stops within 150 feet
-   Example: Route A stops at "State & Madison", Route B stops at "State & Monroe" (both within 150ft) → they share "State" street → overlap verified

**Edge-Based Corridor Identification:**

-   Corridors are identified using an **edge-based methodology** that tracks service between consecutive stop clusters
-   For each trip during peak periods, we create directed edges between consecutive clustered stops
-   Edges are aggregated by (from_cluster, to_cluster, direction_id) to calculate combined frequency
-   This captures the actual path of transit service and correctly handles multi-route overlapping segments
-   **Direction-aware**: Trips are always summed separately by direction (e.g., northbound vs southbound) so opposite-direction service never inflates frequency
-   Combined service interval = peak period duration (120 min) ÷ total trips in that direction

**Agency-Specific Processing (Performance Optimization):**

-   To improve performance with large datasets, corridors are processed separately for each agency
-   **Exception**: PACE and CTA are processed together since their routes can overlap in the Chicago suburbs
-   This ensures overlapping routes from different agencies are correctly combined while avoiding expensive cross-agency comparisons where overlap is impossible

**Qualification Criteria:**

-   **Bus Hubs**: Must have **2 or more routes** AND **combined frequency ≤ 15 minutes** in at least one peak period AND **routes must overlap on same street/intersection**
-   **Bus Corridors**: Must have **1 or more routes** AND **direction-aware combined frequency ≤ 15 minutes** in at least one peak period
  - Routes on the same street that individually don't qualify may qualify together if their combined frequency meets the threshold
  - Directions are kept separate (northbound and southbound are not combined)
  - A corridor qualifies if **any direction in any peak period** has combined frequency ≤15 minutes
-   **Rail Hubs**: All rail stations within Illinois automatically qualify (CTA rail, Metra, and Metro St. Louis MetroLink)

**Direction Handling:**

-   If GTFS `direction_id` data is available, routes are analyzed by direction (northbound/southbound, eastbound/westbound)
-   For corridors, when multiple routes overlap on the same street:
  - Trips are summed across routes **within each direction** (e.g., all northbound trips together)
  - This reflects actual transit availability: if you're waiting for a northbound bus, you can take any northbound route
  - Opposite directions are never combined (bi-directional service is kept separate)

**Edge Geometry Creation:**

1. **Stop clustering:** Group nearby stops into clusters representing physical intersections
2. **Edge building:** Create edges between consecutive clusters along each trip
3. **Aggregation:** Sum trips across all routes using each edge, within each direction/peak
4. **Shape extraction:** Use GTFS shapes.txt to create accurate edge geometries following actual vehicle paths
5. **Qualification:** Edges qualify if combined frequency ≤15 minutes in any direction/peak
6. **Buffering:** Qualified edges are buffered by **680 feet** (660 ft = 1/8 mile + 20 ft half street width) and clipped to Illinois boundary

**Key advantages of edge-based methodology:**

- **Accurate frequency**: Counts actual trips between specific locations, not just at stops
- **Natural overlaps**: Multiple routes using the same segment are automatically combined
- **Shape-based geometry**: Edge segments use GTFS shapes for accurate route paths
- **Direction awareness**: Properly separates opposite-direction service

**Example:**
- Route 8: 4 northbound AM trips between Stop A → Stop B
- Route 11: 4 northbound AM trips between Stop A → Stop B (same segment)
- Combined: 8 northbound AM trips → 120 min / 8 trips = **15 minute combined interval** → Qualifies!

Street-width assumptions: 20 feet is a conservative half-width that covers most Illinois street types (local and arterial) per AASHTO guidance.

**Geographic Boundaries and Clipping:**

-   All transit hub and corridor buffers are clipped to Illinois state boundaries
-   This ensures parking mandate relief applies only to areas within Illinois jurisdiction
-   Metro St. Louis transit system operates across the Illinois-Missouri border; only Illinois portions of hub/corridor buffers are included in the analysis
-   Metra serves some stations in Wisconsin (Kenosha County); these are excluded from the analysis (latitude filter at 42.5°)
-   Clipping is performed after buffer creation using the U.S. Census TIGER/Line state boundary for Illinois

### Statutory Interpretation and Ambiguities

Senate Bill 2111 contains three ambiguities that significantly affect which areas qualify for parking minimum relief. Depending on interpretation, **entire cities could be included or excluded from parking reform** under this legislation.

**1. "Intersecting routes"**: The statute refers to "an intersection of 2 or more bus routes" but does not specify whether this means routes that physically cross each other at a road intersection, or whether it includes routes that overlap on the same street without crossing. **Our interpretation**: Routes must serve stops at the same street intersection (verified through spatial clustering within 150 feet combined with street name matching). Alternative interpretations counting parallel routes on the same street would dramatically expand qualifying areas.

**2. "Combined frequency"**: The statute specifies "combined frequency of service interval" but does not define how frequencies from multiple routes should be combined. Should two routes with 30-minute headways be considered as 30-minute combined service, or 15-minute combined service? **Our interpretation**: We sum total trips across all routes during peak periods in a direction-aware manner (e.g., two routes with 30-minute northbound frequency = 4 northbound trips/hour each = 8 northbound trips/hour = 15-minute combined northbound interval). For corridors, routes sharing the same segment qualify together if their combined frequency (within each direction) meets the threshold. Alternative interpretations using averaged frequencies or minimum headways would result in substantially fewer qualifying locations.

**3. Stop clustering and directionality**: The statute does not specify whether routes must share the exact same stop_id to qualify as "intersecting," or whether nearby stops serving the same intersection should be grouped together. It also does not address whether opposite-direction service should be counted separately or combined. **Our interpretation**: We cluster stops within 150 feet that serve the same street intersection, and when GTFS direction_id data is available, we analyze routes by direction to avoid double-counting bidirectional service. Alternative interpretations requiring exact stop_id matches would exclude many legitimate transit hubs where routes serve nearby stops at the same intersection.

**Legal implications**: Municipalities and courts may interpret these provisions differently, leading to variation in the geographic extent of parking minimum relief across Illinois. This analysis represents one reasonable interpretation aligned with the statute's apparent intent to promote transit-oriented development in areas with genuine high-frequency transit service.

### Data Sources

**Transit Data (GTFS Feeds):**

This analysis incorporates **14 transit agencies** covering major urban areas across Illinois:

-   **Chicago Metro Area:**
    -   CTA (Chicago Transit Authority) - Rail + Bus
    -   Pace Suburban Bus
    -   Metra Commuter Rail
-   **Metro-East (Illinois portion):**
    -   Metro St. Louis (MetroLink Light Rail + MetroBus)
        -   Note: Operates in both Missouri and Illinois; only Illinois portions included after geographic clipping
-   **Champaign-Urbana:** CUMTD (Champaign-Urbana Mass Transit District)
-   **Rockford:** RMTD (Rockford Mass Transit District)
-   **Quad Cities:** MetroLINK (Rock Island County Metropolitan Mass Transit District)
-   **Peoria:** CityLink (Greater Peoria Mass Transit District)
-   **Springfield:** SMTD (Sangamon Mass Transit District)
-   **Bloomington-Normal:** Connect Transit
-   **DeKalb:** DeKalb Public Transit
-   **Decatur:** DPTS (Decatur Public Transit System)
-   **Galesburg:** Galesburg Transit
-   **Macomb:** Go West Transit (Macomb McDonough County Public Transportation)

**Geographic Data:**

-   GTFS shapes.txt from all 14 agencies (actual transit route geometry)
-   Illinois state boundary via TIGER/Line (tigris package)

**Counties Covered:**

The analysis includes **14 Illinois counties** across major urban areas: - **Chicago Metro:** Cook, DuPage, Kane, Lake, McHenry, Will - **Metro-East:** St. Clair - **Central Illinois:** Champaign, McLean, Peoria, Sangamon, Macon - **Northern Illinois:** Winnebago, DeKalb, Rock Island - **Western Illinois:** Knox, McDonough
